---
title: "Getting started with popart"
author: "Zhitan He"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
abstract: |
  A reproducible, offline introduction to preparing two input data frames,
  fitting PopART estimators, and inspecting inference and diagnostics.
vignette: |
  %\VignetteIndexEntry{Getting started with popart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE
)
```

# Overview

`popart` combines a randomized-trial sample with an auxiliary sample that
represents the target population. It returns a trial-only estimator and a
trial-plus-auxiliary estimator for the two treatment-arm means, risk
difference, and risk ratio. Highly adaptive lasso (HAL) is used for the
nuisance regressions.

The installed package analyzes data frames supplied by the user. Data
generation, Monte Carlo studies, and the global HAL scheduler remain in the
repository-level `simulation/` directory and are not part of the installed
API.

# Prepare the two input data frames

The trial data must contain:

- binary treatment, response, censoring, and outcome variables;
- the baseline covariates used for nuisance fitting; and
- one row per trial participant.

The auxiliary data need the same baseline covariates and may include a
nonnegative sampling-weight column. They do not need trial treatment,
response, censoring, or outcome variables.

For an applied analysis, first import the files from their actual locations:

```{r import-user-data, eval=FALSE}
trial_data <- read.csv("path/to/trial_data.csv")
auxiliary_data <- read.csv("path/to/auxiliary_data.csv")
```

To keep this vignette offline and independent of installed example files, the
next chunk deterministically recodes R's built-in `mtcars` data. This coding is
only a computational illustration and has no scientific interpretation.

```{r create-illustration}
library(popart)

cars <- datasets::mtcars[
  rep(seq_len(nrow(datasets::mtcars)), 2L),
  ,
  drop = FALSE
]
rownames(cars) <- NULL
row_id <- seq_len(nrow(cars))

trial <- transform(
  cars,
  outcome = as.integer(mpg >= stats::median(mpg)),
  treatment = as.integer(row_id %% 2L == 0L),
  responded = as.integer(row_id %% 5L != 0L),
  censored = as.integer(row_id %% 3L == 0L)
)
trial$censored[trial$responded == 0L] <- NA_integer_
trial$outcome[
  trial$responded == 0L | trial$censored == 1L
] <- NA_integer_

auxiliary <- cars[c("wt", "hp")]

data.frame(
  data_set = c("trial", "auxiliary"),
  observations = c(nrow(trial), nrow(auxiliary)),
  variables = c(ncol(trial), ncol(auxiliary))
)
```

Missing outcomes are permitted only where an outcome is unobserved. A missing
censoring value is permitted when `responded = 0`. Covariates must be numeric
or logical, finite, and present in both data frames.

# Configure and fit the estimators

The reduced HAL settings below keep the documentation example fast. They are
not a universal recommendation for scientific analyses; use prespecified,
application-appropriate settings for real data.

```{r configure-fit}
teaching_control <- popart_control(
  n_cv_folds = 3L,
  n_lambda_values = 5L,
  n_cv_workers = 1L,
  n_fit_workers = 1L,
  random_seed = 20260806L,
  max_degree = 1L,
  num_knots = 1L,
  keep_nuisance_fits = FALSE
)

teaching_control
```

Column roles are supplied explicitly. `treatment_values` is ordered as
control, then treated. If the auxiliary data have no sampling-weight column,
as in this illustration, unit weights are used.

```{r fit-popart}
fit <- fit_popart(
  trial_data = trial,
  auxiliary_data = auxiliary,
  outcome = "outcome",
  treatment = "treatment",
  response = "responded",
  censoring = "censored",
  covariates = c("wt", "hp"),
  treatment_values = c(0, 1),
  control = teaching_control
)

fit
```

# Inspect estimates and uncertainty

The standard extraction methods provide point estimates, covariance matrices,
and Wald confidence intervals.

```{r inspect-results}
as.data.frame(fit)
coef(fit, estimator = "trial_auxiliary")
confint(fit, estimator = "trial_auxiliary")
round(vcov(fit, estimator = "trial_auxiliary"), 5)
```

The fitted object also records sample sizes, nuisance-fit timing,
fitted-probability ranges, and diagnostic messages.

```{r inspect-diagnostics}
summary(fit)
fit$fit_diagnostics
fit$diagnostics$prediction_ranges
fit$diagnostics$messages
```

Diagnostic thresholds flag probabilities near a boundary; they do not
truncate predictions or silently change the estimator.

# Computational controls

The default detects the CPU slots available to R and uses a conservative number
of complete-fit workers:

```{r parallel-controls}
automatic <- popart_control()

# Run up to two complete nuisance fits concurrently.
fit_parallel <- popart_control(n_fit_workers = 2L, n_cv_workers = 1L)

# Or parallelize three CV folds while complete fits remain serial.
cv_parallel <- popart_control(n_fit_workers = 1L, n_cv_workers = 3L)

automatic
fit_parallel
cv_parallel
```

`n_fit_workers` controls how many of the four complete nuisance fits may run at
once. Automatic mode uses
`max(1, min(4, available_cpu_slots - 2))`. `n_cv_workers` controls workers
for the folds inside a single HAL fit and defaults to one. If CV parallelism is
requested while fit workers are automatic, complete fits are run serially.

Creating a control object does not start workers. Workers are started only when
the object is passed to `fit_popart()`. A worker is an operating-system process
scheduled over available logical CPU slots, not a guaranteed dedicated physical
core. Use only one parallel level at a time because worker processes also share
memory and other system resources.

With the default `keep_nuisance_fits = FALSE`, HAL still performs the same CV
and selects the same model, but the completed fit does not retain its large
`cv.glmnet` object. This reduces result size and worker-transfer overhead; it
does not remove the underlying `glmnet` calculations or reduce peak memory while
an individual fit is running. Set `keep_nuisance_fits = TRUE` only when the
complete fitted nuisance models are needed for later inspection.

# Frequently asked questions

## Does `fit_popart()` read CSV files?

No. Import CSV, database, or other files first, then pass the resulting data
frames to `fit_popart()`.

## Why are two estimators returned?

The trial-only estimator is a randomized-trial comparator. The
trial-plus-auxiliary estimator also uses the auxiliary population's covariate
distribution.

## Is the simulation workflow installed with the package?

No. It remains in the source repository for method evaluation and teaching.
Applied analyses need only the installed package and their two input data
frames.

# References

- Richardson, B. D., Shook-Sa, B. E., and Hudgens, M. G. (n.d.). Causal
  Inference from Cluster-Randomized Trials with Differential Nonresponse.
  Unpublished manuscript submitted to *Biometrics*.

- Hejazi, N. S., Coyle, J. R., and van der Laan, M. J. (2020). `hal9001`:
  Scalable highly adaptive lasso regression in R. *Journal of Open Source
  Software*, 5(53), 2526. doi:
  [10.21105/joss.02526](https://doi.org/10.21105/joss.02526).
