---
title: "Getting started with popart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with popart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Analysis inputs

`popart` combines information from two data frames. The trial data contain the
randomized treatment, response and censoring indicators, outcome, and baseline
covariates. The auxiliary data represent the target population and need only
the same baseline covariates plus an optional sampling weight.

The package includes a small synthetic pair of CSV files so that the expected
layout is concrete. These files are teaching fixtures, not real participant
data or simulation results.

```{r read-data}
library(popart)

trial <- read.csv(
  system.file("extdata", "example_trial.csv", package = "popart")
)
auxiliary <- read.csv(
  system.file("extdata", "example_auxiliary.csv", package = "popart")
)

names(trial)
names(auxiliary)
```

Blank `outcome` values are allowed where the outcome is not observed. A blank
`censored` value is allowed when `responded` is zero. Among respondents,
`censored = 1` means the outcome is censored. Covariates must be numeric or
logical and observed in both data frames.

## Fit the estimators

Column roles are named explicitly, so applied data do not need to use the
example column names.

```{r fit}
fit <- fit_popart(
  trial_data = trial,
  auxiliary_data = auxiliary,
  outcome = "outcome",
  treatment = "treatment",
  response = "responded",
  censoring = "censored",
  covariates = c(
    "baseline_risk",
    "baseline_binary",
    "baseline_score_1",
    "baseline_score_2"
  ),
  auxiliary_weight = "survey_weight",
  treatment_values = c(0, 1)
)
```

`treatment_values` is ordered as control, then treated. If the auxiliary sample
has no sampling weight, omit `auxiliary_weight`; unit weights will be used.

## Inspect results

```{r results}
fit
summary(fit)
coef(fit, estimator = "trial_auxiliary")
vcov(fit, estimator = "trial_auxiliary")
confint(fit, estimator = "trial_auxiliary")
```

The result includes arm-specific means, a risk difference, and a risk ratio for
both the trial-only comparator and the trial-plus-auxiliary analysis. It also
contains nuisance-fit timing and diagnostics. Diagnostics flag potentially
unstable fitted probabilities; they do not automatically change or truncate
the estimates.

## Control computation

Serial execution is the default. A machine with additional available capacity
can run complete nuisance fits concurrently:

```{r controls}
control <- popart_control(
  n_fit_workers = 2,
  n_cv_workers = 1,
  keep_nuisance_fits = FALSE
)

fit_parallel <- fit_popart(
  trial_data = trial,
  auxiliary_data = auxiliary,
  outcome = "outcome",
  treatment = "treatment",
  response = "responded",
  censoring = "censored",
  covariates = c(
    "baseline_risk",
    "baseline_binary",
    "baseline_score_1",
    "baseline_score_2"
  ),
  auxiliary_weight = "survey_weight",
  control = control
)
```

Alternatively, keep `n_fit_workers = 1` and increase `n_cv_workers` to
parallelize CV folds. The control object rejects nested use of both levels.

## Repository simulation workflow

Data generation and Monte Carlo replication are intentionally outside the
installed package. When working from the source repository, see the
"Repository simulation workflow" section of the root `README.md`. That
workflow is for method evaluation; an applied analysis begins with the two
user-supplied data frames shown above.
