---
title: "Getting started with popart"
author: "Zhitan He"
date: "2026-08-05"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
abstract: |
  This vignette gives a reproducible, offline introduction to fitting PopART
  estimators with randomized-trial data and a target-population auxiliary
  sample. It covers input preparation, estimation, inference, diagnostics,
  and computation controls.
vignette: |
  %\VignetteIndexEntry{Getting started with popart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

`popart` estimates treatment-arm means and treatment contrasts by combining a
randomized-trial sample with an auxiliary sample representing the target
population. It reports a trial-only estimator and a trial-plus-auxiliary
estimator. Highly adaptive lasso (HAL) nuisance models are used for outcome,
censoring, and trial-selection mechanisms.

The main workflow is:

1. import the trial and auxiliary files as two data frames;
2. check their variable roles and coding;
3. create an optional computation-control object;
4. call `fit_popart()`; and
5. inspect estimates, confidence intervals, and diagnostics.

The package analyzes supplied data. It does not generate study data or run a
Monte Carlo experiment.

# Prepare the two input data sets

## Required variable roles

The trial data contain:

- a binary randomized treatment;
- a binary response indicator;
- a binary censoring indicator among respondents;
- a binary outcome where that outcome is observed; and
- the baseline covariates used by the nuisance models.

The auxiliary data contain the same baseline covariates and may contain an
optional nonnegative target-population weight. They do not need treatment,
response, censoring, or outcome columns.

Missing outcomes are allowed when an outcome is not observed. A missing
censoring indicator is allowed when the response indicator is zero. Covariates
must be numeric or logical, have no missing values, and be available in both
data sets. Categorical covariates should be encoded before fitting.

## Load the vignette data

The next chunk uses small synthetic CSV files installed in `inst/extdata` so
that the vignette can build offline. This path lookup is specific to the
vignette and is **not** how users normally locate their own files.

```{r load-vignette-data}
library(popart)

example_dir <- system.file("extdata", package = "popart")
trial <- read.csv(file.path(example_dir, "example_trial.csv"))
auxiliary <- read.csv(file.path(example_dir, "example_auxiliary.csv"))

data.frame(
  data_set = c("trial", "auxiliary"),
  observations = c(nrow(trial), nrow(auxiliary)),
  variables = c(ncol(trial), ncol(auxiliary))
)
```

The installed files are teaching fixtures, not real participant data and not
results from the package's repository-level simulation workflow.

```{r inspect-vignette-data}
names(trial)
names(auxiliary)

with(trial, table(treatment, responded, useNA = "ifany"))
summary(auxiliary$survey_weight)
```

For a real analysis, users would instead import files from their own machine or
data platform and then pass the resulting data frames to `fit_popart()`:

```{r real-user-import, eval=FALSE}
trial_data <- read.csv("path/to/trial_data.csv")
auxiliary_data <- read.csv("path/to/auxiliary_data.csv")
```

# Configure the illustrative fit

`popart_control()` records both statistical and computational choices. The
following deliberately small HAL basis and penalty grid keep this documentation
example fast. These teaching settings are not a recommendation for a scientific
analysis; use the package defaults or a prespecified, application-appropriate
configuration for real data.

```{r teaching-control}
teaching_control <- popart_control(
  n_cv_folds = 3L,
  n_lambda_values = 10L,
  n_cv_workers = 1L,
  n_fit_workers = 1L,
  random_seed = 20260805L,
  max_degree = 1L,
  num_knots = c(2L, 1L),
  keep_nuisance_fits = FALSE
)

teaching_control
```

Serial execution makes this vignette reproducible across operating systems and
avoids taking over the reader's computer during package installation.

# Fit PopART

Column roles are supplied explicitly, so users do not need to rename their
variables to match the vignette. `treatment_values` is ordered as control, then
treated. If an auxiliary sample has no sampling-weight column, omit
`auxiliary_weight` and unit weights will be used.

```{r fit-popart}
fit <- fit_popart(
  trial_data = trial,
  auxiliary_data = auxiliary,
  outcome = "outcome",
  treatment = "treatment",
  response = "responded",
  censoring = "censored",
  covariates = c(
    "baseline_risk",
    "baseline_binary",
    "baseline_score_1",
    "baseline_score_2"
  ),
  auxiliary_weight = "survey_weight",
  treatment_values = c(0, 1),
  control = teaching_control
)

fit
```

# Inspect estimates and uncertainty

The fitted object reports control and treated means, their risk difference,
and their risk ratio for both estimators. `trial_only` uses the randomized
trial, whereas `trial_auxiliary` also uses the target-population covariate
distribution supplied by the auxiliary data.

```{r inspect-estimates}
as.data.frame(fit)

coef(fit, estimator = "trial_auxiliary")
confint(fit, estimator = "trial_auxiliary")
```

The corresponding covariance matrix is available through the standard
`vcov()` generic.

```{r inspect-covariance}
round(vcov(fit, estimator = "trial_auxiliary"), 5)
```

# Quality control and diagnostics

`summary()` displays sample sizes, both sets of estimates, and any diagnostic
messages. The fit also stores nuisance-model timing and fitted-probability
ranges for more focused review.

```{r diagnostics}
summary(fit)
fit$fit_diagnostics
fit$diagnostics$prediction_ranges
fit$diagnostics$messages
```

Diagnostic thresholds only flag fitted probabilities near a boundary. They do
not truncate predictions or silently alter the estimator. A warning should be
investigated in the context of the data-generating and sampling processes.

# Computational controls

The default is one complete-fit worker and one CV worker. On a machine with
available capacity, a user can create a control object that runs complete
nuisance fits concurrently:

```{r complete-fit-parallel-control}
fit_parallel_control <- popart_control(
  n_fit_workers = 2L,
  n_cv_workers = 1L
)

fit_parallel_control
```

Alternatively, set `n_fit_workers = 1L` and increase `n_cv_workers` to
parallelize cross-validation folds. The two levels cannot both be greater than
one, which prevents accidental nested parallelism. Creating the control object
does not start workers; workers are used only when it is passed to
`fit_popart()`.

# Repository simulation workflow

Data generation, Monte Carlo replication, and the cross-replication global
scheduler are intentionally outside the installed package. When working from a
source checkout, the root `README.md` explains that research workflow. Those
tools evaluate the method; they are separate from the applied workflow in this
vignette.

# Frequently asked questions

## Does `popart` read CSV files directly?

No. `fit_popart()` accepts data frames. Use `read.csv()`, `readr`, a database
client, or another appropriate import tool first. This separation lets the
package work with many storage formats without embedding file-system behavior
in the estimator.

## Why are there two estimators in the output?

The trial-only result is a comparator based on the randomized trial. The
trial-plus-auxiliary result additionally targets the auxiliary population's
covariate distribution. Differences between them can therefore reflect the
information contributed by that target-population sample.

## Can treatment be labeled instead of coded 0 and 1?

Yes. Supply the observed control and treated values, in that order, through
`treatment_values`, for example `c("usual care", "intervention")`.

## Why might my results differ from this vignette?

This vignette uses synthetic data and a reduced HAL configuration solely to
keep package documentation fast. Results depend on the supplied data, variable
coding, auxiliary weights, randomization probability, random seed, and HAL
settings. Record all of these choices for a reproducible analysis.

## Which parallel option should I use?

Choose one level: either multiple complete-fit workers or multiple CV workers.
Start conservatively because logical processors, available memory, and other
workloads differ across computers. Serial settings remain the most portable.

# Acknowledgments

The implementation relies on the open-source R ecosystem, including
`hal9001` for HAL nuisance fitting and `knitr` and `rmarkdown` for this
documentation. The trial-plus-auxiliary AIPW framework follows the unpublished
manuscript by Richardson, Shook-Sa, and Hudgens; the current HAL nuisance-model
implementation is a separate package design choice. We thank the maintainers
and contributors of the open-source projects used here.

# References

- Richardson, B. D., Shook-Sa, B. E., and Hudgens, M. G. (n.d.). Causal
  Inference from Cluster-Randomized Trials with Differential Nonresponse.
  Unpublished manuscript submitted to *Biometrics*.

- Benkeser, D. and van der Laan, M. (2016). The Highly Adaptive Lasso
  Estimator. *2016 IEEE International Conference on Data Science and Advanced
  Analytics*, 689--696. doi: [10.1109/DSAA.2016.93](https://doi.org/10.1109/DSAA.2016.93).

- Hejazi, N. S., Coyle, J. R., and van der Laan, M. J. (2020). `hal9001`:
  Scalable highly adaptive lasso regression in R. *Journal of Open Source
  Software*, 5(53), 2526. doi:
  [10.21105/joss.02526](https://doi.org/10.21105/joss.02526).
